Lesson Name:  Vectors
Course Name:  R Programming
Type:         Coursera
Author:       Nick Carchedi
Organization: JHU Biostatistics
Version:      2.2.0
================================================================

--- &text

The simplest and most common data structure in R is the vector. 

--- &text

Vectors come in two different flavors: atomic vectors and lists. An atomic vector contains exactly one data type, whereas a list may contain multiple data types. We'll explore atomic vectors further before we get to lists.

--- &text

In previous lessons, we dealt entirely with numeric vectors, which are one type of atomic vector. Other types of atomic vectors include logical, character, integer, and complex. In this lesson, we'll take a closer look at logical and character vectors.

--- &text

Logical vectors can contain the values TRUE, FALSE, and NA (for 'not available'). These values are generated as the result of logical 'conditions'. Let's experiment with some simple conditions.

--- &cmd_question

First, create a numeric vector num_vect that contains the values 0.5, 55, -10, and 6.

```{r}
num_vect <- c(0.5, 55, -10, 6)
```

*** .hint
Recall that the c() function is used for creating a vector. If you forget how to use it, use ?c to access the help file. Don't forget to assign the result to a new variable called num_vect.

--- &cmd_question

Now, create a variable called tf that gets the result of num_vect < 1, which is read as 'num_vect is less than 1'.

```{r}
tf <- num_vect < 1
```

*** .hint
Use tf <- num_vect < 1 to assign the result of num_vect < 1 to a variable called tf.

--- &mult_question

What do you think tf will look like?

_1. a vector of 4 logical values_
2. a single logical value

*** .hint
Remember our lesson on vector arithmetic? The theme was that R performs many operations on an element-by-element basis. We called these 'vectorized' operations.

--- &cmd_question

Print the contents of tf now.

```{r}
tf
```

*** .hint
Just type tf and press Enter to view the its value.

--- &text

The statement num_vect < 1 is a condition and tf tells us whether each corresponding element of our numeric vector num_vect satisfies this condition.

--- &text

The first element of num_vect is 0.5, which is less than 1 and therefore the statement 0.5 < 1 is TRUE. The second element of num_vect is 55, which is greater than 1, so the statement 55 < 1 is FALSE. The same logic applies for the third and forth elements.

--- &cmd_question

Let's try another. Type num_vect >= 6 without assigning the result to a new variable.

```{r}
num_vect >= 6
```

*** .hint
Don't create a new variable on this one. We just want to see the result of num_vect >= 6. Make sure there is no space between the `>` and `=` symbols. Think of them as one symbol representing 'greater than or equal to'.

--- &text

This time, we are asking whether each individual element of num_vect is greater than OR equal to 6. Since only 55 and 6 are greater than or equal to 6, the second and forth elements of the result are TRUE and the first and third elements are FALSE.

--- &text

The `<` and `>=` symbols in these examples are called 'logical operators'. Other logical operators include `>`, `<=`, `==` for exact equality, and `!=` for inequality.

--- &text

If we have two logical expressions, A and B, we can ask whether at least one is TRUE with A | B (logical 'or' a.k.a. 'union') or whether they are both TRUE with A & B (logical 'and' a.k.a. 'intersection'). Lastly, !A is the negation of A and is TRUE when A is FALSE and vice versa.

--- &text

It's a good idea to spend some time playing around with various combinations of these logical operators until you get comfortable with their use. We'll do a few examples here to get you started.

--- &text

Try your best to predict the result of each of the following statements. You can use pencil and paper to work them out if it's helpful. If you get stuck, just guess and you've got a 50% chance of getting the right answer!

--- &mult_question

(3 > 5) & (4 == 4)

1. TRUE
_2. FALSE_

*** .hint
Break this problem down into 2 pieces. The `&` in the middle states that BOTH pieces are TRUE. Your job is to figure out if that is an accurate statement. If so, the entire condition is TRUE. If not, it's FALSE.

--- &mult_question

(TRUE == TRUE) | (TRUE == FALSE)

_1. TRUE_
2. FALSE

*** .hint
Break this problem down into 2 pieces. The `|` in the middle states that AT LEAST ONE of the pieces is TRUE. Your job is to figure out if that is an accurate statement. If so, the entire condition is TRUE. If not, it's FALSE.

--- &mult_question

((111 >= 111) | !(TRUE)) & ((4 + 1) == 5)

_1. TRUE_
2. FALSE

*** .hint
This is a tricky one. Remember that the `!` symbol negates whatever comes after it. There's also an 'order of operations' going on here. Conditions that are enclosed within parentheses should be evaluated first. Then, work your way outwards.

--- &text

Don't worry if you found these to be tricky. They're supposed to be. Working with logical statements in R takes practice, but your efforts will be rewarded in future lessons (e.g. subsetting and control structures).

--- &text

Character vectors are also very common in R. Double quotes are used to distinguish character objects, as in the following example.

--- &cmd_question

Create a character vector that contains the following words: "My", "name", "is". Remember to enclose each word in it's own set of double quotes, so that R knows they are character strings. Store the vector in a variable called my_char.

```{r}
my_char <- c("My", "name", "is")
```

*** .hint
Type my_char <- c("My", "name", "is") to create a new variable called my_char that contains a character vector of length 3. Make sure that the commas separating the words are OUTSIDE of the double quotes, or else R thinks the commas are part of the words.

--- &cmd_question

Print the contents of my_char to see what it looks like.

```{r}
my_char
```

*** .hint
Type my_char to view its contents.

--- &text

Right now, my_char is a character vector of length 3. Let's say we want to join the elements of my_char together into one continuous character string (i.e. a character vector of length 1). We can do this using the paste() function.

--- &cmd_question

Type paste(my_char, collapse = " ") now. Make sure there's a space between the double quotes in the `collapse` argument. You'll see why in a second.

```{r}
paste(my_char, collapse = " ")
```

*** .hint
Use paste(my_char, collapse = " ") to collapse the words in the vector so they almost form a sentence. There should be a single space between the double quotes in the `collapse` argument so that there are single spaces separating the words.

--- &text

The `collapse` argument to the paste() function tells R that when we join together the elements of the my_char character vector, we'd like to separate them with single spaces.

--- &text

It seems that we're missing something.... Ah, yes! Your name!

--- &cmd_question

To add (or 'concatenate') your name to the end of my_char, use the c() function like this: c(my_char, "your_name_here"). Place your name in double quotes where I've put "your_name_here". Try it now, storing the result in a new variable called my_name.

```{r}
my_name <- c(my_char, "Swirl")
```

*** .ans_tests
var_is_a("character", "my_name"); expr_uses_func("c"); expr_creates_var("my_name"); val_has_length(4)

*** .hint
Tack your name on to the end of the my_char vector using the c() function.  Be sure to assign the result to a new variable called my_name. If your name was "Swirl", you would type my_name <- c(my_char, "Swirl").

--- &cmd_question

Take a look at the contents of my_name.

```{r}
my_name
```

*** .hint
Type my_name and press Enter to view its contents.

--- &cmd_question

Now, use the paste() function once more to join the words in my_name together into a single character string. Don't forget to say collapse = " "!

```{r}
paste(my_name, collapse = " ")
```

*** .hint
Use paste(my_name, collapse = " ") to join all four words together, separated by single spaces.

--- &text

In this example, we used the paste() function to collapse the elements of a single character vector. paste() can also be used to join the elements of multiple character vectors.

--- &cmd_question

In the simplest case, we can join two character vectors that are each of length 1 (i.e. join two words). Try paste("Hello", "world!", sep = " "), where the `sep` argument tells R that we want to separate the joined elements with a single space.

```{r}
paste("Hello", "world!", sep = " ")
```

*** .hint
Enter paste("Hello", "world!", sep = " ") to join the two words "Hello" and "world", separated by a single space. There should be a single space between the double quotes in the `sep` argument to the paste() function.

--- &cmd_question

For a slightly more complicated example, we can join two vectors, each of length 3. Use paste() to join the numeric vector 1:3 (recall this is the same as c(1, 2, 3)) with the character vector c("X", "Y", "Z"). This time, use sep = "" to leave no space between the joined elements.

```{r}
paste(1:3, c("X", "Y", "Z"), sep = "")
```

*** .hint
Use paste(1:3, c("X", "Y", "Z"), sep = "") to see what happens when we join two vectors of equal length using paste().

--- &text

What do you think will happen if our vectors are of different length? (Hint: we talked about this in a previous lesson.)

--- &cmd_question

Vector recycling! Try paste(LETTERS, 1:4, sep = "-"), where LETTERS is a predefined variable in R containing a character vector of all 26 letters in the English alphabet.

```{r}
paste(LETTERS, 1:4, sep = "-")
```

*** .hint
Type paste(LETTERS, 1:4, sep = "-") to see how R recycles the vector 1:4 to  match the length of LETTERS. Notice we are using `-` as our separator this time instead of a single space.

--- &text

Since the character vector LETTERS is longer than the numeric vector 1:4, R simply recycles, or repeats, 1:4 until it matches the length of LETTERS.

--- &text

Also worth noting is that the numeric vector 1:4 gets 'coerced' into a character vector by the paste() function. 

--- &text

We'll discuss coercion in another lesson, but all it really means that the numbers 1, 2, 3, and 4 in the output above are no longer numbers to R, but rather characters "1", "2", "3", and "4".
