Lesson Name:  Basic Building Blocks
Course Name:  R Programming
Type:    	    Coursera
Author:       Nick Carchedi
Organization: JHU Biostatistics
Version:      2.2.0
================================================================

--- &text

In this lesson, we will explore some basic building blocks of the R programming language.

--- &text

If at any point you'd like more information on a particular topic related to R, you can type help.start() at the prompt, which will open a menu of resources (either within RStudio or your default web browser, depending on your setup). Alternatively, a simple web search often yields the answer you're looking for.

--- &cmd_question

In it's simplest form, R can be used as an interactive calculator. Type 5 + 7 and press Enter.

```{r}
5 + 7
```

*** .hint
Type 5 + 7 and press Enter.

--- &text

R simply prints the result of 12 by default. However, R is a programming language and often the reason we use a programming language as opposed to a calculator is to automate some process or avoid unnecessary repetition.

--- &text

In this case, we may want to use our result from above in a second calculation. Instead of retyping 5 + 7 every time we need it, we can just create a new variable that stores the result.

--- &text

The way you assign a value to a variable in R is by using the assignment operator, which is just a 'less than' symbol followed by a 'minus' sign. It looks like this: <-

--- &text

Think of the assignment operator as an arrow. You are assigning the value on the right side of the arrow to the variable name on the left side of the arrow.

--- &cmd_question

To assign the result of 5 + 7 to a new variable called x, you type x <- 5 + 7. This can be read as 'x gets 5 plus 7'. Give it a try now.

```{r}
x <- 5 + 7
```

*** .hint
Type x <- 5 + 7. It's important to include a single space on each side of the assignment operator, but do NOT put a space between the `<` and the `-` that form the arrow.

--- &text

You'll notice that R did not print the result of 12 this time. When you use the assignment operator, R assumes that you don't want to see the result immediately, but rather that you intend to use the result for something else later on.

--- &cmd_question

To view the contents of the variable x, just type x and press Enter. Try it now.

```{r}
x
```

*** .hint
Type x and press Enter. This will automatically print the value of x.

--- &cmd_question

Now, store the result of x - 3 in a new variable called y.

```{r}
y <- x - 3
```

*** .hint
Type y <- x - 3 and press Enter. Whether you use x - 3 or x-3 is personal preference, but it's good habit to include a space on either side of the assignment operator.

--- &cmd_question

What is the value of y? Type y to find out.

```{r}
y
```

*** .hint
Type y and press Enter.

--- &text

Now, let's create a small collection of numbers called a vector. Any object that contains data is called a data structure and numeric vectors are the simplest type of data structure in R. In fact, even a single number is considered a vector of length one.

--- &cmd_question

The easiest way to create a vector is with the c() function, which stands for 'concatenate' or 'combine'. To create a vector containing the numbers 1.1, 9, and 3.14, type c(1.1, 9, 3.14). Try it now and store the result in a variable called z.

```{r}
z <- c(1.1, 9, 3.14)
```

*** .hint
Inputting z <- c(1.1, 9, 3.14) will assign the vector (1.1, 9, 3.14) to a new variable called z. Including single spaces after the commas in the vector is not required, but helps make your code less cluttered and more readable.

--- &cmd_question

Anytime you have questions about a particular function, you can access R's built-in help files via the `?` command. For example, if you want more information on the c() function, type ?c without the parentheses that normally follow a function name. Give it a try.

```{r}
?c
```

*** .hint
Type ?c and press Enter. This will bring up the help file for the c() function.

--- &cmd_question

Type z to view it's contents. Notice that there are no commas separating the values in the output.

```{r}
z
```

*** .hint
Type z and press Enter to view its contents.

--- &cmd_question

You can combine vectors to make a new vector. Create a new vector that contains z, 555, then z again in that order. Don't assign this vector to a new variable, so that we can just see the result immediately.

```{r}
c(z, 555, z)
```

*** .hint
Type c(z, 555, z). Don't create a new variable. We just want to view the result.

--- &cmd_question

Numeric vectors can be used in arithmetic expressions. Type the following to see what happens: z * 2 + 100.

```{r}
z * 2 + 100
```

*** .hint
Enter z * 2 + 100, without assigning the result to a variable. The `*` symbol is used for multiplication and shares a key with the number 8 on many keyboards.

--- &text

First, R multiplied each of the three elements in z by 2. Then it added 100 to each element to get the result you see above.

--- &text

Other common arithmetic operators are `+`, `-`, `/`, and `^` (where x^2 means 'x squared'). To take the square root, use the sqrt() function and to take the absolute value, use the abs() function.

--- &cmd_question

Take the square root of z - 1 and assign it to a new variable called mySqrt.

```{r}
mySqrt <- sqrt(z - 1)
```

*** .hint
Variable names are case-sensitive and you are creating a variable called mySqrt, so make sure the letter S is capitalized and everything else is lowercase.

--- &mult_question

Before we view the contents of the mySqrt variable, what do you think it contains?

_1. a vector of length 3_
2. a single number (i.e a vector of length 1)
3. a vector of length 0 (i.e. an empty vector)

*** .hint
Think about how R handled the other 'vectorized' operations: element-by-element.

--- &cmd_question

Print the contents of mySqrt.

```{r}
mySqrt
```

*** .hint
Just type mySqrt and press Enter to view its value.

--- &text

As you may have guessed, R first subtracted 1 from each element of z, then took the square root of each element. This leaves you with a vector of the same length as the original vector z.

--- &cmd_question

Now, create a new variable called myDiv that gets the value of z divided by mySqrt.

```{r}
myDiv <- z / mySqrt
```

*** .hint
Enter myDiv <- z / mySqrt. The spaces on either side of the `/` sign are not required, but can often improve readability by making code appear less cluttered. In the end, it's personal preference.

--- &mult_question

Which statement do you think is true?

_1. The first element of myDiv is equal to the first element of z divided by the first element of mySqrt, and so on..._
2. myDiv is a single number (i.e a vector of length 1)
3. myDiv is undefined

*** .hint
Think about how R handled the other 'vectorized' operations like `+` and `*`.

--- &cmd_question

Go ahead and print the contents of myDiv.

```{r}
myDiv
```

*** .hint
Type myDiv and press Enter to see its contents.

--- &text

When given two vectors of the same length, R simply performs the specified arithmetic operation (`+`, `-`, `*`, etc.) element-by-element. If the vectors are of different lengths, R 'recycles' the shorter vector until it is the same length as the longer vector.

--- &text

When we did z * 2 + 100 in our earlier example, z was a vector of length 3, but technically 2 and 100 are each vectors of length 1.

--- &text

Behind the scenes, R is 'recycling' the 2 to make a vector of 2s and the 100 to make a vector of 100s. In other words, when you ask R to compute z * 2 + 100, what it really computes is this: z * c(2, 2, 2) + c(100, 100, 100).

--- &cmd_question

To see another example of how this vector 'recycling' works, try adding c(1, 2, 3, 4) and c(0, 10). Don't worry about saving the result in a new variable.

```{r}
c(1, 2, 3, 4) + c(0, 10)
```

*** .hint
Enter c(1, 2, 3, 4) + c(0, 10) in the console to see how R adds two vectors of different length. Don't assign the result to a variable.

--- &text

If the length of the shorter vector does not divide evenly into the length of the longer vector, R will still apply the 'recycling' method, but will throw a warning to let you know something fishy might be going on.

--- &cmd_question

Try c(1, 2, 3, 4) + c(0, 10, 100) for an example.

```{r}
c(1, 2, 3, 4) + c(0, 10, 100)
```

*** .hint
Type c(1, 2, 3, 4) + c(0, 10, 100) to see how R handles adding two vectors, when the shorter vector's length does not divide evenly into the longer vector's length. Don't worry about assigning the result to a variable.