Lesson Name:  Sequences of Numbers
Course Name:  R Programming
Type:         Coursera
Author:       Nick Carchedi
Organization: JHU Biostatistics
Version:      2.2.0
================================================================

--- &text

In this lesson, you'll learn how to create sequences of numbers in R.

--- &cmd_question

The simplest way to create a sequence of numbers in R is by using the `:` operator. Type 1:20 to see how it works.

```{r}
1:20
```

*** .hint
Type 1:20 with no spaces. Make sure to use a regular colon and NOT a semi-colon.

--- &cmd_question

That gave us every integer between (and including) 1 and 20. We could also use it to create a sequence of real numbers. For example, try pi:10.

```{r}
pi:10
```

*** .hint
Enter pi:10 and see what happens. pi is a predefined constant in R that takes on the value 3.1415....

--- &text

The result is a vector of real numbers starting with pi (3.142...) and increasing in increments of 1. The upper limit of 10 is never reached, since the next number in our sequence would be greater than 10.

--- &cmd_question

What happens if we do this: 15:1? Give it a try to find out.

```{r}
15:1
```

*** .hint
Type 15:1 and press Enter.

--- &text

It counted backwards in increments of 1! It's unlikely we'd want this behavior, but nonetheless it's good to know how it could happen.

--- &text

Remember that if you have questions about a particular R function, you can access its documentation with a question mark followed by the function name: ?function_name_here. However, in the case of an operator like the colon used above, you must enclose the symbol in backticks like this: ?`:`. (NOTE: The backtick (`) key is generally located in the top left corner of a keyboard, above the Tab key.)

--- &cmd_question

Pull up the documentation for `:` now.

```{r}
?`:`
```

*** .hint
In order to view the documentation for a symbol like the colon operator, you have to use backticks. This is so R knows you are not attempting to use the symbol in the command. Here's what it looks like: ?`:`. Don't forgot the question mark out front.

--- &text

Often, we'll desire more control over a sequence we're creating than what the `:` operator gives us. The seq() function serves this purpose.

--- &cmd_question

The most basic use of seq() does exactly the same thing as the `:` operator. Try seq(1, 20) to see this.

```{r}
seq(1, 20)
```

*** .hint
Type seq(1, 20) and press Enter. The space after the comma is not required, but is recommended as it tends to make your code appear less cluttered.

--- &cmd_question

This gives us the same output as 1:20. However, let's say that instead we want a vector of numbers ranging from 0 to 10, incremented by 0.5. seq(0, 10, by=0.5) does just that. Try it out.

```{r}
seq(0, 10, by=0.5)
```

*** .hint
You are still using the seq() function here, but this time with an extra argument that tells R you want to increment your sequence by 0.5. Try seq(0, 10, by=0.5).

--- &cmd_question

Or maybe we don't care what the increment is and we just want a sequence of 30 numbers between 5 and 10. seq(5, 10, length=30) does the trick. Give it shot now and store the result in a new variable called my_seq.

```{r}
my_seq <- seq(5, 10, length=30)
```

*** .hint
You're using the same function here, but changing its arguments for different results. Be sure to store the result in a new variable called my_seq, like this: my_seq <- seq(5, 10, length=30).

--- &cmd_question

To confirm that my_seq has length 30, we can use the length() function. Try it now.

```{r}
length(my_seq)
```

*** .hint
Use length(my_seq) to view the length of the my_seq variable.

--- &text

Let's pretend we don't know the length of my_seq, but we want to generate a sequence of integers from 1 to N, where N represents the length of the my_seq vector. In other words, we want a new vector (1, 2, 3, ...) that is the same length as my_seq.

--- &cmd_question

There are several ways we could do this. One possibility is to combine the `:` operator and the length() function like this: 1:length(my_seq). Give that a try.

```{r}
1:length(my_seq)
```

*** .hint
Try 1:length(my_seq), without assigning the result to a new variable.

--- &cmd_question

Another option is to use seq(along = my_seq). Give that a try.

```{r}
seq(along = my_seq)
```

*** .hint
Type seq(along = my_seq). Notice we're using the same seq() function, but with a different argument called `along`. This tells R that you want a sequence 'along' the sequence my_seq. Remember you can type ?seq if you have questions about the seq() function.

--- &cmd_question

However, as is the case with many common tasks, R has a separate built-in function for this purpose called seq_along(). Type seq_along(my_seq) to see it in action.

```{r}
seq_along(my_seq)
```

*** .hint
The seq_along() function is a variation of the seq() function. Try seq_along(my_seq) to get the same result yet another way.

--- &text

There are often several approaches to solving the same problem, particularly in R. Simple approaches that involve less typing are generally best. It's also important for your code to be readable, so that you and others can figure out what's going on without too much hassle.

--- &text

If R has a built-in function for a particular task, it's likely that function is highly optimized for that purpose and is your best option. As you become a more advanced R programmer, you'll design your own functions to perform tasks when there are no better options. We'll explore writing your own functions in future lessons.

--- &text

One more function related to creating sequences of numbers is rep(), which stands for 'replicate'. Let's look at a few uses.

--- &cmd_question

If we're interesting in creating a vector that contains 40 zeros, we can use rep(0, times = 40). Try it out.

```{r}
rep(0, times = 40)
```

*** .hint
Type rep(0, times = 40) to make a vector containing 40 zeros.

--- &cmd_question

If instead we want our vector to contain 10 repetitions of the vector (0, 1, 2), we can do rep(c(0, 1, 2), times = 10). Go ahead.

```{r}
rep(c(0, 1, 2), times = 10)
```

*** .hint
Try rep(c(0, 1, 2), times = 10) for a different variation on the same theme. Be sure to use the c() function to tell R that the numbers 0, 1, and 2 make up a vector.

--- &cmd_question

Finally, let's say that rather than repeating the vector (0, 1, 2) over and over again, we want our vector to contain 10 zeros, then 10 ones, then 10 twos. We can do this with the `each` argument. Try rep(c(0, 1, 2), each = 10).

```{r}
rep(c(0, 1, 2), each = 10)
```

*** .hint
Type rep(c(0, 1, 2), each = 10) to see how the `each` argument to the rep() function alters its behavior slightly.
